# Seeking Help

<hr class="double">

## Reading Help files

`R`, and every package, provide help files for functions. To search for help on a specific function that is in a package loaded into your namespace (your interactive `R` session):

```{r help1, eval=FALSE}
?function_name

help(function_name)

function_name  # Highlight the function name without parenthesis and press F1
```

This will load up a help page in `RStudio` (or as plain text in `R` by itself).

Each help page is broken down into sections:

+  Description: An extended description of what the function does.
+  Usage: The arguments of the function and their default values (if any).
+  Arguments: An explanation of the input each argument is expecting.
+  Details: Any important details to be aware of.
+  Value: The data the function returns.
+  See Also: Any related functions you might find useful.
+  Examples: Some examples for how to use the function.

Different functions might have different sections, but these are the main ones you should be aware of.

<hr class="small">

## Special Operators

To seek help on special operators, use quotes:

```{r help2, eval=FALSE}
?"+"
```

<hr class="small">

## Getting help on packages

Many packages come with "vignettes": tutorials and extended example documentation. Without any arguments, `vignette()` will list all vignettes for all installed packages; `vignette(package="package-name")` will list all available vignettes for`package-name`, and `vignette("vignette-name")` will open the specified vignette.

If a package doesn't have any vignettes, you can usually find help by typing
`help("package-name")`.

<hr class="small">

## When you kind of remember the function

If you're not sure what package a function is in, or how it's specifically spelled you can do a fuzzy search:

```{r help3, eval=FALSE}
??function_name
```

<hr class="small">

## When you have no idea where to begin

If you don't know what function or package you need to use [CRAN Task Views](http://cran.at.r-project.org/web/views) is a specially maintained list of packages grouped into fields. This can be a good starting point.

<hr class="small">

## When your code doesn't work: seeking help from your peers

If you're having trouble using a function, 9 times out of 10, the answers you are seeking have already been answered on [Stack Overflow](http://stackoverflow.com/). You can search using the `[r]` tag.

If you can't find the answer, there are a few useful functions to help you ask a question from your peers:

```{r help4, eval=FALSE}
?dput
```

Will dump the data you're working with into a format so that it can be copy and pasted by anyone else into their `R` session.

```{r help5}
sessionInfo()
```

Will print out your current version of `R`, as well as any packages you have loaded. This can be useful for others to help reproduce and debug your issue.

```{r Challenge_8}
###################
### Challenge 8 ###
###################

# Look at the help for the `paste` function. You'll need to use this later.  What is the difference between the `sep` and `collapse` arguments?
```

<hr class="small">

## Other ports of call

+  [Quick R](http://www.statmethods.net/)
+  [RStudio cheat sheets](http://www.rstudio.com/resources/cheatsheets/)
+  [Cookbook for R](http://www.cookbook-r.com/)

<hr class="small">
